Symbolic Weighted Language Models and\\ Parsing over Infinite Alphabets

### 2021-05-25 main-V2.tex
V2 : section parsing : Def. complete & proof

Let $\Semiring$ be a commutative semiring, 
$\Sigma$ and $\Delta$ be alphabets called respectively \emph{input} and \emph{output}, %{alphabets}, 
and $\bar\Phi = \< \Phi_\Sigma, \Phi_\Delta, \Phi_{\Sigma, \Delta}>$ be an $\Semiring$-label theory.

@@

\section{SW Visibly Pushdown Automata}
\label{section:SWVPA}\label{sec:SWVPA}
The following model generalizes Symbolic VPA~\cite{dAntonyAlur14SVPDA}
from Boolean semirings to arbitrary semiring weight domains.

@@

\subsection{Properties}
Like VPA and symbolic VPA, 
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$, $\Semiring$ and $\bar\Phi$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as in the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.

@@

\subsection{Trees and Nested-Words}
\label{sec:trees}
The hierarchical structure of nested-words, defined with the \emph{call} and \emph{return} markup symbols  
suggest a correspondence with trees. 
The lifting of this correspondence to languages, of tree automata and VPA,
has been discussed in~\cite{AlurMadhusudan09nested}, 
and~\cite{Caralp12VPAmult} for the weighted case.
In this section, we describe a correspondence between the symbolic-weighted extensions
of tree automata and VPA.

Let $\Omega$ be a countable ranked alphabet, such that 
every symbol $a \in \Omega$ has a rank 
$\rank(a) \in [0..M]$ where $M$ is a fixed natural number.
We denote by $\Omega_k$ the subset of all symbols $a$ of $\Omega$
with $\rank(a) = k$, where $0 \leq k \leq M$, 
and $\Omega_{>0} = \Omega \setminus \Omega_0$.
%
\noindent 
The free $\Omega$-algebra of finite, ordered, 
$\Omega$-labeled trees is denoted by $\T(\Omega)$.
It is the smallest set such that  $\Omega_0 \subset \T(\Omega)$
and for all $1 \leq k \leq M$, all $a \in \Omega_k$, 
and all $t_1, \ldots, t_k \in \T(\Omega)$, $a(t_1, \ldots, t_k) \in \T(\Omega)$.
%
% tree = single node (leave) labeled with a symbol of $a \in \Omegai$
% (such a tree is simply denoted by $a$)
% or the composition, denoted by $b(t_1,\ldots, t_n$) of a node labeled with $b$
% and $n$ subtrees $t_1$,\ldots, $t_n$.
%
Let us assume a  commutative semiring $\Semiring$ 
and a label theory $\bar{\Phi}$ over $\Semiring$ 
containing one set $\Phi_{\Omega_k}$ for each $k \in [0..M]$.

@@

\begin{lemma}  \label{lem:SWTA}
For all \SWTA $A$ over~$\Omega$, $\Semiring$ commutative, and $\Phi$,
there exists an effectively constructible \SWVPA $A'$ over 
$\hat\Omega$, $\Semiring$ and $\hat\Phi$ 
such that for all $t \in \T(\Omega)$, $A'\bigl(\lin(t)\bigr) = A(t)$.
\end{lemma} 
% 
\begin{proof}
Let $A = \< Q, \init, \bar{\wei} >$ where $\bar{\wei}$ is presented as above by a function
    %
We build 
$A' = \< Q', P', \init', \bar{\wei}', \final' >$,
%computing over $\hat\Omega = \Omegai \uplus \Omegac \uplus \Omegar$,
where $Q' = \bigcup_{k=0}^{M} Q^k$ is the set of sequences of state symbols of $A$, 
of length at most $M$, including the empty sequence denoted by $\epsilon$, 
and where $P' = Q'$ and $\bar\wei$ is defined by:

\[
\begin{array}{ll}
\weii(\bar{q}, {\call{b}}, \bar{p}, a, \bar{q} q') = \wei(q', a, \epsilon) & 
\mathrm{for~all~} b \in \Omega_{>0}, \bar{p}\in P', a \in \Omega_0\\
%
\weiei(\bar{q}, a, \bar{q} q') = \wei(q', a, \epsilon) & 
\mathrm{for~all~} a \in \Omega_0\\
%
\weic(\bar{q}, {\call{b}}, \bar{p}, \call{b}, \epsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0},  \bar{p}\in P', b \in \Omega_{>0}\\
%
\weiec(\bar{q}, \call{b}, \epsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0}\\
%
%\weir: Q \times \Omegac \times P \times \Sigmar \times Q \to \Semiring & 
\weir(\bar{q}, {\call{b}}, \bar{p}, {\return{b}}, \bar{p} q') = \wei(q', b, \bar{q}) & 
\mathrm{for~all~}  b \in \Omega_{>0}, \bar{p}\in P'\\
%
%\weie: Q \times \Omegar \times Q \to \Semiring & 
\weier(\bar{p}, {\return{b}}, \bar{q}) = \zero &
\mathrm{for~all~}  b \in \Omega_{>0}\\
\end{array}      
\]
\noindent
It is sufficient to consider in $Q'$ only the prefixes of 
sequences in transition with a non-null weight.
\qed\end{proof}

\section{Symbolic Weighted Parsing}
\label{sec:parsing}
Let us now use the models and results of the former sections in %order to define 
an approach to the problem of parsing over infinite alphabet. %appropriate
%
Besides considering infinitely many possible of input symbols, 
handled with suitable language formalisms, 
this approach extends conventional parsing 
by computing a derivation tree modulo 
a generic distance between words,  
defined by a SW transducer given in input.
This enables considering finer word relationships than strict equality 
as in the conventional parsing approach, 
opening possibilities of quantitative analysis via this method.

@@

Let $\Sigma$ be a countable input alphabet 
and $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$ an output countable alphabet,
let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative, bounded, complete 
and total semiring 
and let  $(\Phi_\epsilon, \Phic, \Phir, \Phicr)$ be a label theory over $\Semiring$, 
assumed computable and $k$-convex for some fixed $k$.

\noindent
Assuming given in input:
\begin{itemize}
\item a \SWT $T$ over $\Sigma$,  $\Omega$, and $\Semiring$,
defining a measure %between words 
$T: \Sigma^* \times \Omega^* \to \Semiring$,
\item a \SWVPA $A$ over $\Omega$, and $\Semiring$, defining a series of nested words
      $A : \Omega^* \to \Semiring$,
\item an input word $s \in \Sigma^*$,
\end{itemize}
the problem of \emph{symbolic weighted parsing} is to find 
a nested word $t \in \Omega$ minimizing
\( T(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$, 
\ie such that %Hence, it is the problem of finding 
\( T(s, t) \otimes A(t) = 
  \displaystyle\bigoplus_{t' \in \Omega^*} T(s, t') \otimes A(t') \).

\noindent 
Therefore, it is the problem of optimizing a measure 
called the \emph{edit-distance between $s$ and $A$} in~\cite{Mohri03EDWA}.
%
The input language can also be expressed as a \SWTA, or, 
as a particular case, as a weighted context-free grammar, 
converted in turn into a \SWVPA following Lemma~\ref{lem:SWTA}.
%
In the case of finite alphabets, 
the problem of searching, in a WTA language, 
for the best parse tree for a given,  
sometimes referred as~\emph{weighted parsing}
(see \cite{Goodman99SemiringParsing,MorbitzVogler19weighted-parsing} **more general problems**)
is a particular case of SW parsing.
%
Indeed, it corresponds to the case where $T$
accepts only the pairs $\<s, t>$ such that 
$s$ is the projection of $t$ on $\Omegai$. 
This can be done with a single state $q$ and 
with transition rules of the form:
\begin{description}
\item[] $\wei(q, \epsilon, a, q) = \one$ for all $a \in \Omegac \cup \Omegar$,
\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Omegai$,
\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Omegai$, $a \neq b$.
\end{description}

@@

\begin{proposition}
The problem of Symbolic Weighted  parsing 
can be solved in PTIME in the size of the input \SWT, \SWVPA (or \SWTA) and input word, 
and the computation time of the functions of the label theory.
\end{proposition}
%
%\begin{proof}
%Bar-Hillel construction
%\qed  
%\end{proof}

%\input{transcription}

>>>>

Let $\Semiring$ be a commutative semiring, 
$\Sigma$ and $\Delta$ be alphabets called respectively \emph{input} and \emph{output}, %{alphabets}, 
and $\bar\Phi = \< \Phi_\Sigma, \Phi_\Delta, \Phi_{\Sigma, \Delta}>$ be a label theory over $\Semiring$.

@@

\section{SW Visibly Pushdown Automata}
\label{section:SWVPA}\label{sec:SWVPA}
The model presented in this section generalizes Symbolic VPA~\cite{dAntonyAlur14SVPDA}
from Boolean semirings to arbitrary semiring weight domains.
It will compute on nested words over infinite alphabets, 
associating to every such word a weight value. 
Nest words are able to describe structures of labeled trees, 
and in the context of parsing, they will be useful to 
represent parse trees.

@@

\subsection{Properties}
Like VPA and symbolic VPA, 
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$, $\Semiring$ and $\bar\Phi$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as in the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.

@@

\subsection{Nested-Words and Parse-Trees}
\label{sec:trees}
The hierarchical structure of nested-words, defined with the \emph{call} and \emph{return} markup symbols  
suggest a correspondence with trees. 
The lifting of this correspondence to languages, of tree automata and VPA,
has been discussed in~\cite{AlurMadhusudan09nested}, 
and~\cite{Caralp12VPAmult} for the weighted case.
In this section, we describe a correspondence between the symbolic-weighted extensions
of tree automata and VPA.

Let $\Omega$ be a countable ranked alphabet, such that 
every symbol $a \in \Omega$ has a rank 
$\rank(a) \in [0..M]$ where $M$ is a fixed natural number.
We denote by $\Omega_k$ the subset of all symbols $a$ of $\Omega$
with $\rank(a) = k$, where $0 \leq k \leq M$, 
and $\Omega_{>0} = \Omega \setminus \Omega_0$.
%
\noindent 
The free $\Omega$-algebra of finite, ordered, 
$\Omega$-labeled trees is denoted by $\T(\Omega)$.
It is the smallest set such that  $\Omega_0 \subset \T(\Omega)$
and for all $1 \leq k \leq M$, all $a \in \Omega_k$, 
and all $t_1, \ldots, t_k \in \T(\Omega)$, $a(t_1, \ldots, t_k) \in \T(\Omega)$.
%
% tree = single node (leave) labeled with a symbol of $a \in \Omegai$
% (such a tree is simply denoted by $a$)
% or the composition, denoted by $b(t_1,\ldots, t_n$) of a node labeled with $b$
% and $n$ subtrees $t_1$,\ldots, $t_n$.
%
Let us assume a commutative semiring~$\Semiring$ 
and a label theory~$\bar{\Phi}$ over~$\Semiring$ 
containing one set~$\Phi_{\Omega_k}$ for each $k \in [0..M]$.

@@

\begin{lemma}\label{lem:SWTA}
For all \SWTA $A$ over~$\Omega$, $\Semiring$ commutative, and $\Phi$,
there exists an effectively constructible \SWVPA $A'$ over 
$\hat\Omega$, $\Semiring$ and $\hat\Phi$ 
such that for all $t \in \T(\Omega)$, $A'\bigl(\lin(t)\bigr) = A(t)$.
\end{lemma} 
% 
\begin{proof}
Let $A = \< Q, \init, \bar{\wei} >$ where $\bar{\wei}$ is presented as above by a function
    %
We build 
$A' = \< Q', P', \init', \bar{\wei}', \final' >$,
%computing over $\hat\Omega = \Omegai \uplus \Omegac \uplus \Omegar$,
where $Q' = \bigcup_{k=0}^{M} Q^k$ is the set of sequences of state symbols of $A$, 
of length at most $M$, including the empty sequence denoted by $\epsilon$, 
and where $P' = Q'$ and $\bar\wei$ is defined by:

\[
\begin{array}{ll}
\weii(\bar{q}, {\call{b}}, \bar{p}, a, \bar{q} q') = \wei(q', a, \epsilon) & 
\mathrm{for~all~} b \in \Omega_{>0}, \bar{p}\in P', a \in \Omega_0\\
%
\weiei(\bar{q}, a, \bar{q} q') = \wei(q', a, \epsilon) & 
\mathrm{for~all~} a \in \Omega_0\\
%
\weic(\bar{q}, {\call{b}}, \bar{p}, \call{b'}, \epsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0},  \bar{p}\in P', b, b' \in \Omega_{>0}\\
%
\weiec(\bar{q}, \call{b}, \epsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0}\\
%
%\weir: Q \times \Omegac \times P \times \Sigmar \times Q \to \Semiring & 
\weir(\bar{q}, {\call{b}}, \bar{p}, {\return{b}}, \bar{p} q') = \wei(q', b, \bar{q}) & 
\mathrm{for~all~}  b \in \Omega_{>0}, \bar{p}\in P'\\
%
%\weie: Q \times \Omegar \times Q \to \Semiring & 
\weier(\bar{p}, {\return{b}}, \bar{q}) = \zero &
\mathrm{for~all~}  b \in \Omega_{>0}\\
\end{array}      
\]
\noindent
It is sufficient to consider in $Q'$ only the prefixes of 
sequences in transition with a non-null weight.
\qed\end{proof}

\section{Symbolic Weighted Parsing}
\label{sec:parsing}
Let us now apply the models and results of the previous sections %in order to define 
to the problem of parsing over infinite alphabet. %appropriate
%
Besides considering infinitely many possible of input symbols, 
handled with suitable language formalisms, 
this approach extends conventional parsing and weighted parsing
by computing a derivation tree modulo 
a generic distance between words,  
defined by a SW transducer given in input.
This enables considering finer word relationships than strict equality 
as in the conventional parsing approach, 
opening possibilities of quantitative analysis via this method.

@@

Let $\Sigma$ be a countable input alphabet 
and let $\Omega$ be a countable output ranked alphabet, 
with maximal rank value $M$, 
and $\hat\Omega = \Omegai \uplus \Omegac \uplus \Omegar$
be the alphabet with nesting symbols, associated like Section~\ref{sec:trees}, 
for the linearization of trees of $\T(\Omega)$
(remember that $\Omegai = \Omega_0$).
%
Let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative, bounded, complete 
and total semiring 
and let $\bar\Phi$ be a label theory over $\Semiring$
containing
$\Phi_\Sigma$, $\Phi_{\Sigma, \Omegai}$, as well as
$\Phii$, $\Phic$, $\Phir$, $\Phicr$
(following the notations of Section~\ref{se:SWVPA-def}).
It is moreover assumed computable and $k$-convex for some fixed $k$.

\noindent
Let us assume given the following input:
\begin{itemize}
\item a \SWT $T$ over $\Sigma$, $\Omegai$, $\Semiring$, and $\bar\Phi$, 
defining a measure %between words 
$T: \Sigma^* \times \Omegai^* \to \Semiring$,

\item a \SWTA $A$ over $\Omega$, $\Semiring$, and $\bar\Phi$,
defining a measure $A: \T(\Omega) \to \Semiring$,
%a \SWVPA $A$ over $\Omega$, and $\Semiring$, defining a series of nested words
%      $A : \Omega^* \to \Semiring$,
\item an input word $s \in \Sigma^*$.
\end{itemize}
%
As explained in Section~\ref{sec:trees}, 
$A = \< Q, \init, \bar{\wei} >$ can be seen as a weighted regular tree grammar, 
that generates (weighted) trees by replacement of a state symbol $q_0$ (non-terminal), 
by a tree $a(q_1,\ldots, q_k)$, where $k = \rank(a)$.
A replacement rule $q_0 \to a(q_1,\ldots, q_k)$, 
of weight $\wei(q_0, a, q_1 \ldots q_k) \in \Semiring$ 
according to Definition~\ref{def:SWTA}, 
corresponds to the production rule $q_0 := a(q_1,\ldots, q_k)$ of a weighted CF grammar,
with set non-terminal symbols $Q$ and set of terminal symbols $\Omega_0$.
This actually is a slight generalization of CFG since 
each such production rule is labelled by  a symbol of $\Omega_{>0}$, 
hence parse trees %derivation trees
are trees of $\T(\Omega)$.
%
Another (more original) generalization is that the set of terminal symbols 
$\Omega_0$ may be infinite.

\noindent 
We extend the measure defined by $T$ to 
$d : \Sigma^* \times \T(\Omega) \to \Semiring$ as follows.
Given a word $w \in {\hat\Omega}^*$, the projection of $w$ onto $\Omegai$,
denoted $w|_\Omegai$,
is the word of $\Omegai^*$ obtained from $w$ by removing all symbols 
in $\hat\Omega \setminus \Omegai$.
Using this notation
and the tree linearization operator defined in Section~\ref{sec:trees}, 
$d$ is defined by:
\begin{equation}
d(s, t) = T\bigl(s, \lin(t)|_\Omegai \bigr)  \mathrm{~for~} s \in \Sigma^*, t\in \T(\Omega)
\end{equation}

\noindent 
\emph{Symbolic weighted parsing} is the problem, 
given the above input, 
to find a tree $t \in \T(\Omega)$ %nested word $t \in \Omega$ 
minimizing \( d(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$, 
\ie such that: %Hence, it is the problem of finding 
%
\begin{equation}\label{eq:distance-lang}
d(s, t) \otimes A(t) = \displaystyle\bigoplus_{t' \in \T(\Omega)} d(s, t') \otimes A(t') 
\end{equation}
%
The measure expressed in~\eqref{eq:distance-lang} is called
called edit-distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%
%The input language can also be expressed as a \SWTA, or, 
%as a particular case, as a weighted context-free grammar, 
%converted in turn into a \SWVPA following Lemma~\ref{lem:SWTA}.
%
The problem of searching, in a WTA language, 
the best parse tree matching a given input,  
sometimes referred as~\emph{weighted parsing}
corresponds to SW parsing in the case of finite alphabets
and when the transducer $T$ characterizes identity
see \eg~\cite{Goodman99SemiringParsing} 
and~\cite{MorbitzVogler19weighted-parsing} for a more general 
weighted parsing framework.
%
%Indeed, it corresponds to the case where $T$
%accepts only the pairs $\<s, t>$ such that 
%$s$ is the projection of $t$ on $\Omegai$. 
%This can be done with a single state $q$ and 
%with transition rules of the form:
%\begin{description}
%\item[] $\wei(q, \epsilon, a, q) = \one$ for all $a \in \Omegac \cup \Omegar$,
%\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Omegai$,
%\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Omegai$, $a \neq b$.
%\end{description}

@@

\begin{proposition}
The problem of Symbolic Weighted  parsing 
can be solved in PTIME in the size of the input \SWT $T$, \SWTA $A$ %\SWVPA (or ) 
and input word $s$, 
and the computation time of the functions of the label theory.
\end{proposition}
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, also called parsing by intersection.

\noindent
We first extend the \SWT $T$ over $\Sigma$, $\Omegai$, $\Semiring$, and $\bar\Phi$, 
into a \SWT $T'$ over $\Sigma$ and $\hat\Omega$ (and the same semiring and label theory),
such that for all $s \in \Sigma^*$, and $u \in {\hat\Omega}^*$, 
$T'(s, u) = T(s, u|_{\Omegai})$. The transducer $T'$ simplify skips every symbol 
$b \in {\hat\Omega} \setminus \Omegai$ in input, 
with new transitions of the form $\wei_{01}(q, \epsilon, b, q')$.

\noindent
Then, given an input word $s \in \Sigma^*$, 
we compute the \SWA $A_{T', s}$, using Proposition~\ref{prop:epsilon}.
This automaton is such that for all $t \in \T(\Omega)$, 
$A_{T', s}\bigl(\lin(t)\bigr) 
 = T'\bigl(s, \lin(t)\bigr) 
 = T'\bigl(s, \lin(t)|_{\Omegai}\bigr) 
 = d(s, t)$.

\noindent
Next, we convert the input \SWTA $A$ over $\Omega$
into a \SWVPA $A'$ over $\hat\Omega$, using Lemma~\ref{lem:SWTA}, 
and we compute the \SWVPA $A_{T', s} \otimes A'$, 
using Proposition~\ref{prop:SWVPA-product}.

\noindent
It remains to compute a best nested-word $w \in {\hat\Omega}^*$ 
using the best-search procedure of Proposition~\ref{th:best-search},
and convert it into a best tree in $\T(\Omega)$ in order to solve SW parsing
for~$T$, $A$ and~$s$.
\qed  
\end{proof}

\subsection{Application to Automated Music Transcription}
%\input{transcription}
